---
title: "Classic PAPRIKA Mode"
author: "fairpaprika"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classic PAPRIKA Mode}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

The `fairpaprika` package implements both the original **Classic PAPRIKA** algorithm (Hansen & Ombler, 2008) and an enhanced **Full Mode** with active learning and uncertainty quantification.

This vignette covers **Classic PAPRIKA Mode**, which provides:

- Systematic two-criterion pairwise questions
- Dominance-based elimination
- Transitive closure for efficiency
- Adaptive outcome-based stopping
- Real-time Pareto set tracking

## Classic vs. Full Mode

| Feature | Classic Mode | Full Mode |
|---------|-------------|-----------|
| Question selection | Systematic pairs | Active learning (EIG) |
| Weights | Deterministic (LP) | Probabilistic (sampling) |
| Stopping | Profile resolution | Convergence + budget |
| Uncertainty | None | Full posterior |
| Fairness metrics | Basic | Enhanced (Gini, voice) |

## Basic Usage

```{r}
library(fairpaprika)

# Define decision criteria
domains <- list(
    Efficacy = c("Low", "Medium", "High"),
    SideEffects = c("Severe", "Moderate", "Mild"),
    Cost = c("Expensive", "Moderate", "Cheap")
)

# Create engine in classic mode
settings <- list(
    mode = "classic",
    min_q = 3, # Minimum questions before stopping
    max_q = 50 # Maximum question budget
)

eng <- engine_create(domains, settings = settings)
```

## Setting Treatment Profiles

Classic mode shines when ranking specific alternatives:

```{r}
# Define therapy options
therapies <- data.frame(
    Efficacy = c("High", "Medium", "Low"),
    SideEffects = c("Mild", "Moderate", "Severe"),
    Cost = c("Expensive", "Moderate", "Cheap"),
    stringsAsFactors = FALSE
)
rownames(therapies) <- c("Therapy_A", "Therapy_B", "Therapy_C")

eng <- engine_set_profiles(eng, therapies)
```

## Elicitation Loop

```{r}
while (!engine_done(eng)) {
    # Get next question
    nxt <- engine_next_question(eng)
    if (is.null(nxt$question)) break

    # Present to user (in real app, show via UI)
    cat("Question: Which do you prefer?\n")
    cat("  A:", paste(names(nxt$question$a), nxt$question$a, sep = "=", collapse = ", "), "\n")
    cat("  B:", paste(names(nxt$question$b), nxt$question$b, sep = "=", collapse = ", "), "\n")

    # Get user response
    pref <- readline("Enter A, B, or E (equal): ")

    # Record decision
    eng <- nxt$engine
    eng <- engine_add_decision(eng, pref = pref)

    # Show Pareto set update
    cat("\n", paprika_pareto_summary(eng), "\n\n")
}
```

## Adaptive Stopping

Classic PAPRIKA stops when **all profile pairs are resolved**:

```{r}
# Check if stopping criterion is met
if (paprika_all_profiles_resolved(eng)) {
    cat("All profiles can be strictly ordered - stopping!\n")
}
```

This typically requires **far fewer** questions than the maximum budget:

```{r}
# Example: 5 profiles, 12 possible pairs
# Questions asked: 6
# Reduction: 50% (asked only 6 instead of 12)
```

## Computing Results

```{r}
eng <- engine_compute(eng)

# View weights
print(eng$weights)

# View winner probabilities (deterministic in classic mode)
print(eng$diagnostics$winner_probabilities)
```

## Enhanced Features

### 1. Real-Time Pareto Tracking

Track which alternatives remain competitive:

```{r}
# After each decision
pareto_summary <- paprika_pareto_summary(eng)
# Returns: "Potentially optimal: 2/5 profiles
#           Therapy_A, Therapy_C
#           Just eliminated: Therapy_E"
```

### 2. Dominated Profile Pruning

Automatically excludes inferior options:

```{r}
dominated <- eng$diagnostics$dominated_profiles
dominance_pairs <- eng$diagnostics$dominance_pairs

# dominance_pairs shows: Therapy_A dominates Therapy_B
```

### 3. Efficiency Diagnostics

Measure question efficiency:

```{r}
diag <- paprika_diagnostics(eng)

# Questions asked vs. total possible
print(diag$efficiency$question_efficiency) # e.g., 0.25 = 75% reduction

# Preferences implied by transitivity
print(diag$efficiency$implied_preferences) # e.g., 15 inferred from 5 asked
```

### 4. Inconsistency Detection

Detect contradictions in responses:

```{r}
if (diag$inconsistencies$has_inconsistencies) {
    cat("Warning: Inconsistent responses detected\n")
    print(diag$inconsistencies$inconsistency_details)
}
```

## Comparison to Traditional PAPRIKA

**Original PAPRIKA (1995-2008):**
- Manual question enumeration
- No real-time feedback
- Post-hoc dominance checking

**fairpaprika Classic Mode:**
- ✅ All original features
- ✅ **Plus**: Real-time Pareto tracking
- ✅ **Plus**: Adaptive stopping
- ✅ **Plus**: Enhanced diagnostics
- ✅ **Plus**: Inconsistency detection

## When to Use Classic Mode

**Use Classic Mode when:**
- You want deterministic, explainable weights
- Ranking a small set of known alternatives (< 10)
- Computational simplicity is important
- No need for uncertainty quantification

**Use Full Mode when:**
- You need uncertainty estimates (confidence intervals)
- Large alternative sets or complex interactions
- Procedural fairness metrics are important (Gini, voice)
- Active learning for maximum efficiency

## Complete Example

```{r, eval=FALSE}
library(fairpaprika)

# Setup
domains <- list(
    Quality = c("Poor", "Good", "Excellent"),
    Price = c("High", "Medium", "Low")
)

products <- data.frame(
    Quality = c("Excellent", "Good", "Poor"),
    Price = c("High", "Medium", "Low"),
    stringsAsFactors = FALSE
)
rownames(products) <- c("Premium", "Standard", "Budget")

# Create and run
eng <- engine_create(domains, settings = list(mode = "classic"))
eng <- engine_set_profiles(eng, products)

# Simulate consistent user
set.seed(123)
while (!engine_done(eng)) {
    nxt <- engine_next_question(eng)
    if (is.null(nxt$question)) break

    # Prefer higher quality
    a_score <- match(nxt$question$a$Quality, domains$Quality)
    b_score <- match(nxt$question$b$Quality, domains$Quality)
    pref <- if (a_score > b_score) "A" else if (b_score > a_score) "B" else "E"

    eng <- nxt$engine
    eng <- engine_add_decision(eng, pref = pref)
}

# Results
eng <- engine_compute(eng)
cat(sprintf(
    "Winner: %s\n",
    names(which.max(eng$diagnostics$winner_probabilities))
))

# Diagnostics
diag <- paprika_diagnostics(eng)
cat(diag$summary)
```

## References

Hansen, P., & Ombler, F. (2008). A new method for scoring additive multi-attribute value models using pairwise rankings of alternatives. *Journal of Multi-Criteria Decision Analysis*, 15(3-4), 87-107.

## See Also

- `vignette("fairpaprika")` - Full mode with active learning
- `vignette("healthcare-sdm")` - Clinical decision support applications
- `?paprika_diagnostics` - Enhanced diagnostic functions
