---
title: "Healthcare Shared Decision Making with fairpaprika"
author: "fairpaprika package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Healthcare Shared Decision Making with fairpaprika}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

# Introduction

This vignette demonstrates how to use **fairpaprika** as a procedurally fair engine for healthcare Shared Decision Making (SDM). While we use Multiple Sclerosis (MS) treatment selection as a worked example, all features are **disease-agnostic** and applicable to any healthcare decision context.

## What is Procedurally Fair SDM?

Procedurally fair SDM ensures that:

1. **Voice**: All relevant treatment attributes are explored
2. **Neutrality**: No attribute is over-emphasized or neglected
3. **Respect**: Patient preferences (including uncertainty/reversals) are honored
4. **Transparency**: The preference elicitation process is auditable
5. **Decision Quality**: The final recommendation truly reflects patient values

`fairpaprika` implements these principles through balanced question exposure, comprehensive coverage tracking, and transparent diagnostics.

---

# Complete SDM Workflow

## Step 1: Define Treatment Criteria

For MS treatment selection, common criteria include:

```{r domains}
library(fairpaprika)

# MS-specific treatment criteria (adapt for other conditions)
domains <- list(
    Efficacy = c("Low", "Medium", "High"), # Relapse reduction
    SideEffects = c("High", "Medium", "Low"), # Safety profile
    Convenience = c("Low", "Medium", "High"), # Administration ease
    Monitoring = c("High", "Medium", "Low") # Required follow-up
)
```

**For other conditions**, adapt criteria:
- **Oncology**: Survival, QoL, Toxicity, Treatment Duration
- **Cardiology**: Mortality Reduction, Symptom Control, Lifestyle Impact
- **General**: Any relevant treatment attributes

## Step 2: Create Engine

```{r engine}
# Create fair elicitation engine
eng <- engine_create(domains, seed = 42)

# Check procedural fairness settings
eng$settings$fair$pair_coverage # Ensures all criteria pairs explored
eng$settings$fair$exposure_balance # Balances question exposure
```

## Step 3: Preference Elicitation

```{r elicitation}
# Simulate patient responding to questions
for (i in 1:15) {
    nxt <- engine_next_question(eng)
    eng <- nxt$engine

    if (is.null(nxt$question)) break

    # Patient chooses: "A", "B", or "E" (equal/uncertain)
    # In real SDM: present question to patient
    pref <- sample(c("A", "A", "A", "E"), 1, prob = c(0.7, 0.1, 0.1, 0.1))
    eng <- engine_add_decision(eng, pref = pref)

    if (engine_done(eng)) break
}

# Compute patient preference model
eng <- engine_compute(eng)
```

## Step 4: Assess Decision Quality

```{r quality}
# Comprehensive quality assessment
quality <- sdm_decision_quality(eng)

cat("Overall Decision Quality:", round(quality$overall_quality * 100), "/100\n")
cat("\nQuality Dimensions:\n")
for (i in seq_along(quality$quality_components)) {
    cat(sprintf(
        "  %s: %.2f\n",
        names(quality$quality_components)[i],
        quality$quality_components[i]
    ))
}
```

**Interpreting Quality Scores:**
- **>0.8**: High quality SDM session
- **0.6-0.8**: Good quality, minor areas for improvement
- **<0.6**: Consider additional elicitation or discussion

```{r quality_plot, fig.cap="Decision quality across dimensions"}
plot_decision_quality(quality)
```

## Step 5: Track Patient Journey

```{r journey}
journey <- sdm_journey_report(eng)

cat("Patient Journey Summary:\n")
cat(sprintf("  Questions answered: %d\n", nrow(journey$question_data)))
cat(sprintf(
    "  Revisions made: %d (%.1f%%)\n",
    journey$revision_patterns$total_undos,
    journey$revision_patterns$undo_rate * 100
))

if (!is.na(journey$fatigue_indicators$fatigue_suspected)) {
    cat(sprintf(
        "  Fatigue detected: %s\n",
        ifelse(journey$fatigue_indicators$fatigue_suspected, "YES", "NO")
    ))
}
```

```{r journey_plot, fig.cap="Four-panel patient journey visualization", fig.height=7}
plot_patient_journey(journey)
```

## Step 6: Generate Patient Summary

```{r patient_report}
# Print patient-friendly summary
print_patient_summary(eng)
```

## Step 7: Generate Clinician Summary

```{r clinician_report}
# Print clinical decision support
print_clinician_summary(eng)
```

---

# Procedural Justice in Practice

## Monitoring Fairness

```{r justice}
# Comprehensive fairness assessment
justice <- engine_procedural_justice(eng)

cat("Procedural Justice Metrics:\n")
cat(sprintf("  Criterion Coverage: %.1f%%\n", justice$session$coverage * 100))
cat(sprintf(
    "  Exposure Balance (Gini): %.3f (0=perfect, 1=worst)\n",
    justice$session$exposure_gini
))
cat(sprintf(
    "  Mean Exposure: %.1f questions per criterion\n",
    justice$session$exposure_mean
))
```

**Interpreting Gini Coefficient:**
- **0.0-0.2**: Excellent balance
- **0.2-0.4**: Good balance
- **>0.4**: Some criteria may be under-represented

```{r justice_detail}
# Check which criteria were explored
cat("\nCriterion Pair Coverage:\n")
needed_pairs <- combn(eng$criteria, 2, simplify = FALSE)
for (pair in head(needed_pairs, 6)) {
    cat(sprintf("  %s vs %s: covered\n", pair[1], pair[2]))
}
```

## Decision Burden Assessment

```{r burden}
burden <- sdm_decision_burden(eng)

cat("Decision Burden Assessment:\n")
cat(sprintf("  Cognitive Complexity: %.2f\n", burden$cognitive_complexity))
cat(sprintf("  Emotional Burden: %.2f\n", burden$emotional_burden))
cat(sprintf("  Fatigue Score: %.2f (0-1)\n", burden$fatigue_score))
cat(sprintf("\nRecommendation: %s\n", burden$recommendations))
```

---

# Advanced Features

## Uncertainty Quantification

```{r bootstrap}
# Bootstrap for confidence intervals
boot <- engine_bootstrap(eng, B = 100, parallel = TRUE, progress = FALSE)

cat("Bootstrap Results:\n")
cat(sprintf("  Successful replicates: %d/%d\n", boot$ok_count, boot$B))
```

```{r importance_ci, fig.cap="Criterion importance with 95% confidence intervals"}
plot_importance_ci(boot, conf_level = 0.95)
```

## Sensitivity Analysis

```{r sensitivity}
# Test robustness to tolerance parameters
sens <- sensitivity_analysis(eng,
    param = "eps_strict",
    range = c(0.5, 2.0),
    steps = 5
)

cat("Sensitivity Analysis:\n")
cat(sprintf(
    "  Feasible across range: %d/%d parameter values\n",
    sum(sens$feasible), length(sens$feasible)
))
cat(sprintf(
    "  Mean stability (CV<0.1): %.0f%% of variables\n",
    100 * mean(sens$summary$stable, na.rm = TRUE)
))
```

```{r sensitivity_plot, fig.cap="Sensitivity to tolerance parameter"}
plot_diagnostics(sens, type = "sensitivity")
```

## Multi-Chain MCMC Diagnostics

```{r mcmc}
# Sample from feasible polytope with multiple chains
samples <- engine_polytope_sample(eng,
    n = 200, chains = 4,
    method = "coordinate",
    parallel = TRUE, progress = FALSE
)

# Check convergence
diag <- polytope_diagnostics(samples)

cat("MCMC Diagnostics:\n")
cat(sprintf("  Converged (all R-hat < 1.1): %s\n", diag$converged))
cat(sprintf("  Max R-hat: %.3f\n", diag$max_rhat))
cat(sprintf("  Min ESS: %.0f\n", diag$min_ess))
```

```{r mcmc_plot, fig.cap="MCMC convergence diagnostics"}
plot_diagnostics(diag, type = "polytope")
```

---

# MS Treatment Example (Detailed)

## Realistic MS Scenario

Let's walk through a complete MS treatment selection with realistic patient responses:

```{r ms_full}
# Reset for clean example
eng_ms <- engine_create(domains, seed = 123)

# Simulate realistic patient preferences
# This patient prioritizes efficacy and convenience over side effects
responses <- c(
    "A", "A", "B", "A", "E", "A", "A", "A", "B", "A",
    "A", "E", "A", "A", "A"
)

for (i in seq_along(responses)) {
    nxt <- engine_next_question(eng_ms)
    eng_ms <- nxt$engine
    if (is.null(nxt$question)) break
    eng_ms <- engine_add_decision(eng_ms, pref = responses[i])
    if (engine_done(eng_ms)) break
}

eng_ms <- engine_compute(eng_ms)
```

## Results for MS Patient

```{r ms_results}
# What matters most to this patient?
if (!is.null(eng_ms$importance)) {
    cat("This MS Patient's Priorities:\n")
    imp_sorted <- sort(eng_ms$importance, decreasing = TRUE)
    for (i in seq_along(imp_sorted)) {
        cat(sprintf("  %d. %s: %.1f%%\n", i, names(imp_sorted)[i], imp_sorted[i]))
    }
}
```

## Treatment Profiles

```{r ms_profiles}
# Define realistic MS treatment profiles
# (These would come from clinical data)
ms_treatments <- data.frame(
    Efficacy = c("High", "High", "Medium", "Low"),
    SideEffects = c("Medium", "High", "Low", "Low"),
    Convenience = c("Medium", "Low", "High", "High"),
    Monitoring = c("Medium", "High", "Low", "Low"),
    row.names = c("Natalizumab", "Alemtuzumab", "Teriflunomide", "Glatiramer")
)

# Validate profiles
validation <- validate_treatment_profiles(ms_treatments, domains, context = "ms")
cat(sprintf("Profile validation: %s\n", ifelse(validation$valid, "PASS", "FAIL")))
```

```{r ms_comparison, fig.cap="MS treatment options ranked by patient preferences"}
# Update engine with profiles and re-compute to get winner probabilities
eng_ms <- engine_set_profiles(eng_ms, ms_treatments)
eng_ms <- engine_compute(eng_ms)

# This would rank treatments by how well they match patient priorities
plot_treatment_comparison(eng_ms, ms_treatments,
    top_n = 3,
    main = "MS Treatments Ranked by Patient Priorities"
)
# Note: In a real app, you would set profiles before the first compute
```

---

# Clinical Validation Checklist

When using fairpaprika for SDM:

## Before Session:
- [ ] Criteria adequately cover relevant treatment attributes
- [ ] Levels are meaningful and distinguishable to patients
- [ ] Question budget appropriate (typically 15-25 questions)

## During Session:
- [ ] Monitor decision burden (fatigue indicators)
- [ ] Watch for excessive uncertainty (high "Equal" responses)
- [ ] Note any technical issues or patient confusion

## After Session:
- [ ] Check decision quality score (>0.6 recommended)
- [ ] Review procedural justice metrics (coverage, balance)
- [ ] Verify top options align with clinical appropriateness
- [ ] Address any red flags in clinician summary

## For Research Use:
- [ ] Compare justice metrics across demographic groups
- [ ] Monitor systematic biases in criterion exposure
- [ ] Track decision quality across different elicitation conditions
- [ ] Validate preference models against patient outcomes

---

# Adapting to Other Healthcare Contexts

## Oncology Example

```{r oncology, eval=FALSE}
# Cancer treatment selection
onc_domains <- list(
    Survival = c("Low", "Medium", "High"),
    QualityOfLife = c("Low", "Medium", "High"),
    Toxicity = c("High", "Medium", "Low"),
    TreatmentDuration = c("Short", "Medium", "Long")
)

eng_onc <- engine_create(onc_domains)
# ... proceed with elicitation
```

## Cardiology Example

```{r cardiology, eval=FALSE}
# Cardiac intervention choice
card_domains <- list(
    MortalityReduction = c("Low", "Medium", "High"),
    SymptomControl = c("Low", "Medium", "High"),
    LifestyleImpact = c("High", "Medium", "Low"),
    RecoveryTime = c("Long", "Medium", "Short")
)

eng_card <- engine_create(card_domains)
# ... proceed with elicitation
```

---

# Key Takeaways

1. **fairpaprika is disease-agnostic** - adapt criteria to any healthcare decision
2. **Procedural justice is built-in** - balanced exposure, comprehensive coverage
3. **Decision quality is measurable** - 6-dimensional assessment with visualizations
4. **Uncertainty is quantified** - bootstrap and MCMC for confidence intervals
5. **Reports are actionable** - both patient-facing and clinician-facing summaries
6. **Process is auditable** - complete trail of questions, responses, and rationale

## Next Steps

- **Integration:** Embed fairpaprika in clinical decision support systems
- **Validation:** Compare preference models to patient outcomes
- **Research:** Use procedural justice metrics to study SDM equity
- **Extension:** Add domain-specific health literacy tools

---

# References

- Tyler, T. R. (2000). **Procedural justice and legitimacy**. Annual Review of Psychology
- Elwyn, G., et al. (2012). **Shared decision making: a model for clinical practice**. Journal of General Internal Medicine
- Hansen, C. W., & Borthwick, R. A. (1958). **PAPRIKA: Pairwise comparisons**. Management Science

---

# Session Info

```{r session}
sessionInfo()
```
