---
title: "Advanced: Interaction Terms"
author: "fairpaprika package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: Interaction Terms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Interaction Terms in Preference Models

This vignette explains **when and how to use interaction terms** in fairpaprika. Interactions allow non-additive preferences - situations where the value of one criterion depends on another.

> **Note:** Interactions add computational complexity. Start without them and add only if needed.

---

## What Are Interactions?

### Additive Model (Default)

```
utility(Treatment) = w1·Effect + w2·SideEffects + w3·Convenience
```

Each criterion contributes independently.

### Model with Interactions

```
utility(Treatment) = w1·Effect + w2·SideEffects + w3·Convenience
                   + w12·(Effect×SideEffects)
```

The interaction term `w12` captures synergy or trade-offs between criteria.

---

## When to Use Interactions

### Use interactions when:

1. **Synergies exist**
   - "High efficacy is only valuable if side effects are low"
   - "Convenience matters more when monitoring burden is high"

2. **Non-compensatory preferences**
   - "I won't accept high side effects no matter how effective"
   - Strict thresholds or deal-breakers

3. **Model doesn't fit well**
   - High uncertainty in additive model
   - Bootstrap CIs are very wide
   - Slack is needed for feasibility

### Don't use interactions when:

1. **Additive model works well** (most cases!)
2. **Limited decisions** (<15 questions) - insufficient data
3. **Simplicity is priority** - additive models are more interpretable

---

## Enabling Interactions

```{r enable_interactions}
library(fairpaprika)

domains <- list(
    Effect = c("Low", "Medium", "High"),
    SideEffects = c("High", "Medium", "Low")
)

# Enable interaction terms
eng_int <- engine_create(domains,
    settings = list(
        interactions = list(
            enabled = TRUE,
            pairs = list(c("Effect", "SideEffects"))
        )
    )
)
```

**Key settings:**
- `enabled`: Master switch (default: TRUE but no specific pairs)
- `pairs`: List of criterion pairs to model, e.g., `list(c("A", "B"), c("B", "C"))`
- `max_abs`: Bound on interaction coefficients (default: 1.5)

---

## Model Comparison

Compare models with and without interactions:

```{r model_comparison}
# Create both engines
eng_base <- engine_create(domains, seed = 42)

eng_int <- engine_create(domains,
    seed = 42,
    settings = list(
        interactions = list(
            enabled = TRUE,
            pairs = list(c("Effect", "SideEffects"))
        )
    )
)

# Collect same decisions for both
for (i in 1:10) {
    nxt_base <- engine_next_question(eng_base)
    nxt_int <- engine_next_question(eng_int)

    eng_base <- nxt_base$engine
    eng_int <- nxt_int$engine

    if (is.null(nxt_base$question)) break

    # Same responses
    pref <- "A"
    eng_base <- engine_add_decision(eng_base, pref = pref)
    eng_int <- engine_add_decision(eng_int, pref = pref)
}

# Compare models
comp <- model_comparison(eng_base, eng_int, n_samples = 100)

cat("Model Comparison:\n")
cat(sprintf("  Base model feasible: %s\n", comp$feasible[1]))
cat(sprintf("  Interaction model feasible: %s\n", comp$feasible[2]))
cat(sprintf("  Variance ratio (int/base): %.2f\n", comp$variance_ratio))
cat(sprintf("\n%s\n", comp$recommendation))
```

**Interpreting results:**
- **Variance ratio > 1.5**: Interaction model is substantially more flexible
- **Variance ratio < 1.2**: Models are similar - prefer simpler additive
- **Variance ratio 1.2-1.5**: Marginal benefit - use clinical judgment

---

## Types of Interactions

### 1. Conditional Tradeoffs

```{r conditional_example, eval=FALSE}
# Example: "Would you accept Medium Effect with Low SideEffects
#          vs. High Effect with Medium SideEffects?"

# This tests if patient values Effect differently
# depending on SideEffects level
```

**Interpretation:**
- Positive interaction: Criteria synergize (both high is extra valuable)
- Negative interaction: Criteria conflict (trade-off is steep)

### 2. Joint Improvements

```{r joint_example, eval=FALSE}
# Example: "Option A: (Effect:Low, Side:High)
#          vs. Option B: (Effect:Medium, Side:Medium)"

# This tests if simultaneous improvement matters more than sum of parts
```

---

## Computational Costs

Interactions have costs:

| Aspect | Additive | With Interactions |
|--------|----------|-------------------|
| **Questions** | 15-20 | 20-30 |
| **Computation** | O(n) | O(n²) |
| **Interpretability** | High | Moderate |
| **Data needs** | Low | High |

**Rule of thumb:**
- **k criteria** → k(k-1)/2 possible interaction pairs
- Each pair adds ~2-5 questions to test
- Limit to 1-3 most important pairs

---

## Interaction Settings

```{r interaction_settings}
# Full configuration
settings_full <- list(
    interactions = list(
        enabled = TRUE,

        # Which pairs to model (leave empty for automatic)
        pairs = list(c("Effect", "SideEffects")),

        # How many pairs can be active simultaneously
        max_pairs = 2,

        # Maximum fraction of questions that are interaction questions
        max_fraction = 0.25, # 25% max

        # Bound on interaction coefficients
        max_abs = 1.5,

        # When to ask interaction questions
        activate = list(
            winner_entropy = 1.0, # High uncertainty triggers
            slack = TRUE, # Infeasibility triggers
            min_questions = 6 # Wait until enough base questions
        ),

        # Question types
        families = list(
            conditional = TRUE, # Conditional tradeoff questions
            joint = TRUE # Joint improvement questions
        )
    )
)

eng_configured <- engine_create(domains, settings = settings_full)
```

---

## Interpreting Interaction Coefficients

```{r interpret_interactions, eval=FALSE}
# After solving
eng_int <- engine_compute(eng_int)

# Look at weights
print(eng_int$weights)

# Interaction terms have names like "Effect:High×SideEffects:Low"
# Positive: Synergy (both high is extra good)
# Negative: Conflict (trade-off is steep)
# Near zero: Independent (additive model sufficient)
```

**Clinical example:**
```
w(Effect:High) = 0.45
w(SideEffects:Low) = 0.30
w(Effect:High × SideEffects:Low) = 0.15  # POSITIVE synergy

→ Treatment with High Effect + Low SideEffects gets:
  0.45 + 0.30 + 0.15 = 0.90 total utility
  
→ The combination is worth MORE than sum of parts
```

---

## Testing if Interactions are Needed

### Method 1: Bootstrap Uncertainty

```{r bootstrap_test, eval=FALSE}
# Additive model
boot_base <- engine_bootstrap(eng_base, B = 100)

# Check confidence intervals
plot_importance_ci(boot_base)

# If CIs are very wide → might need interactions
```

### Method 2: Slack Detection

```{r slack_test, eval=FALSE}
eng_base <- engine_compute(eng_base)

if (!is.null(eng_base$diagnostics$slack_used)) {
    if (eng_base$diagnostics$slack_used > 0) {
        cat("Slack detected - consider interaction model\n")
    }
}
```

### Method 3: Residual Analysis

```{r residual_test, eval=FALSE}
# Sample polytope and check variance
samples <- engine_polytope_sample(eng_base, n = 500)

# High variance → model uncertainty → maybe interactions help
```

---

## Practical Workflow

```{r workflow, eval=FALSE}
# Step 1: Start with additive model
eng <- engine_create(domains)
# ... collect decisions ...
eng <- engine_compute(eng)

# Step 2: Check if additive is sufficient
quality <- sdm_decision_quality(eng)

if (quality$overall_quality > 0.7) {
    # Additive model is fine!
    cat("Additive model sufficient\n")
} else {
    # Step 3: Try interaction model
    eng_int <- engine_create(domains,
        settings = list(
            interactions = list(
                enabled = TRUE,
                pairs = list(c("Crit1", "Crit2"))
            )
        )
    )

    # Re-elicit or transfer decisions
    # ...

    # Step 4: Compare
    comp <- model_comparison(eng, eng_int)
    print(comp$recommendation)
}
```

---

## Common Pitfalls

### 1. Too Many Interactions

```{r too_many, eval=FALSE}
# BAD: All possible pairs for 5 criteria = 10 pairs!
# Would need 50+ questions

# GOOD: 1-2 most important pairs
pairs <- list(c("Effect", "SideEffects"))
```

### 2. Insufficient Data

```{r insufficient_data, eval=FALSE}
# Need at least 4-5 questions per interaction pair
# With <15 total questions → stick to additive
```

### 3. Over-interpretation

```{r over_interpret, eval=FALSE}
# Small interaction coefficients (<0.05) are likely noise
# Focus on large, clinically meaningful interactions (>0.1)
```

---

## Interaction-Specific Diagnostics

```{r interaction_diagnostics, eval=FALSE}
# Check if interactions were actually used
if (!is.null(eng_int$interactions_pairs_active)) {
    cat("Active interaction pairs:\n")
    print(eng_int$interactions_pairs_active)
}

# Check interaction question count
audit <- eng_int$audit
int_questions <- sum(vapply(
    audit,
    function(x) isTRUE(x$interaction_pair),
    logical(1)
))

cat(sprintf(
    "Interaction questions: %d/%d (%.0f%%)\n",
    int_questions,
    length(audit),
    100 * int_questions / length(audit)
))
```

---

## When Interactions Really Matter

### Healthcare Examples:

1. **MS Treatments:**
   - *Effect × SideEffects*: "Only accept high-efficacy drugs if side effects are manageable"
   
2. **Cancer Treatments:**
   - *Survival × QoL*: "Extra months matter more if quality of life is good"
   
3. **Surgical Options:**
   - *Recovery × Complications*: "Longer recovery acceptable if complication risk is low"

### When NOT to use:

- Routine decisions with clear trade-offs
- Simple preference structures
- Limited patient engagement (short sessions)

---

## Key Takeaways

1. **Start without interactions** - most preferences are additive
2. **Add judiciously** - only 1-3 most important pairs
3. **Test necessity** - compare models, check uncertainty
4. **Increase budget** - need 20-30 questions with interactions
5. **Interpret carefully** - small coefficients may be noise
6. **Document rationale** - explain why interactions were included

**For most SDM applications, the additive model is sufficient and more interpretable!**

---

## References

- Fishburn, P. C. (1967). **Additive utilities with incomplete product sets**. *Operations Research*
- Keeney, R. L., & Raiffa, H. (1993). **Decisions with Multiple Objectives**. Cambridge.

```{r session_info}
sessionInfo()
```
