---
title: "Advanced: Tolerance Parameter Tuning"
author: "fairpaprika package"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced: Tolerance Parameter Tuning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Tolerance Parameters in fairpaprika

This vignette explains the three key tolerance parameters that control the mathematical strictness of preference constraints in fairpaprika. These are **advanced settings** - the defaults work well for most applications.

## The Three Parameters

### 1. `eps_strict` (Default: 1e-3)

**Purpose:** Minimum utility margin for strict preferences ("A preferred to B")

**Mathematical role:**
```
If patient prefers A over B: utility(A) - utility(B) ≥ eps_strict
```

**When to adjust:**
- **Increase** (to 0.01-0.05) if you want stronger preference differentiation
- **Decrease** (to 1e-4) if model becomes infeasible with many decisions
- **Never** set below 1e-6 (numerical precision limits)

**Example:**
```{r eps_strict_example}
library(fairpaprika)

domains <- list(A = c("Low", "High"), B = c("Low", "High"))

# Strict enforcement (larger margin)
eng_strict <- engine_create(domains, settings = list(eps_strict = 0.01))

# Lenient enforcement (smaller margin)
eng_lenient <- engine_create(domains, settings = list(eps_strict = 1e-4))
```

---

### 2. `tau_equal` (Default: 1e-3)

**Purpose:** Tolerance band for indifference responses ("A and B are equal")

**Mathematical role:**
```
If patient says A ~ B: |utility(A) - utility(B)| ≤ tau_equal
```

**When to adjust:**
- **Increase** (to 0.01-0.05) to allow more flexibility in "equal" responses
- **Decrease** (to 1e-4) for stricter enforcement of indifference
- Related to `eps_strict` - typically set to same value

**Example:**
```{r tau_equal_example}
# Wide indifference band
eng_wide <- engine_create(domains, settings = list(tau_equal = 0.05))

# Narrow indifference band
eng_narrow <- engine_create(domains, settings = list(tau_equal = 1e-4))
```

---

### 3. `epsilon_monotone` (Default: 1e-6)

**Purpose:** Minimum increment between adjacent levels within a criterion

**Mathematical role:**
```
For ordered levels (Low < Med < High):
utility(Med) - utility(Low) ≥ epsilon_monotone
utility(High) - utility(Med) ≥ epsilon_monotone
```

**When to adjust:**
- **Rarely changed** - this is a numerical stability parameter
- **Increase** (to 1e-4) only if you want to force meaningful level differences
- **Never** set to zero (would allow degenerate solutions)

**Example:**
```{r epsilon_monotone_example}
# Force larger increments
eng_increments <- engine_create(domains,
    settings = list(epsilon_monotone = 1e-4)
)
```

---

## Sensitivity Analysis

Test how robust your results are to parameter choices:

```{r sensitivity_demo, fig.width=7, fig.height=5}
# Create example engine
eng <- engine_create(domains, seed = 42)

# Collect some decisions
for (i in 1:5) {
    nxt <- engine_next_question(eng)
    eng <- nxt$engine
    if (is.null(nxt$question)) break
    eng <- engine_add_decision(eng, pref = "A")
}

eng <- engine_compute(eng)

# Test sensitivity to eps_strict
sens <- sensitivity_analysis(eng,
    param = "eps_strict",
    range = c(0.1, 5.0), # 10% to 500% of default
    steps = 10
)

# Check results
cat("Feasibility across parameter range:\n")
print(sens$feasible)

cat("\nStability assessment:\n")
print(sens$summary)

# Visualize
plot_diagnostics(sens, type = "sensitivity")
```

**Interpreting results:**
- **CV < 0.1**: Variable is stable (robust to parameter)
- **CV 0.1-0.3**: Moderate sensitivity
- **CV > 0.3**: High sensitivity (results depend on parameter choice)

---

## Decision Tree: When to Adjust Parameters

```
START: Using default parameters (eps_strict=1e-3, tau_equal=1e-3, epsilon_monotone=1e-6)
  |
  ├─ Model is FEASIBLE and results look reasonable
  │  └─ DON'T CHANGE (defaults are working)
  |
  ├─ Model is INFEASIBLE (solve fails)
  │  ├─ Reduce eps_strict to 1e-4
  │  ├─ Reduce tau_equal to 1e-4
  │  └─ If still infeasible: check for contradictory decisions
  |
  ├─ All weights are nearly identical (no differentiation)
  │  └─ Increase eps_strict to 0.01
  |
  └─ Indifference responses seem ignored
     └─ Increase tau_equal to 0.01
```

---

## Common Scenarios

### Scenario 1: Inconsistent Preferences

**Problem:** Patient gave contradictory responses, model uses slack

```{r inconsistent}
# Check slack usage
eng_check <- engine_create(domains)
# ... collect decisions with contradictions ...
# eng_check <- engine_compute(eng_check)

# If slack is used:
# eng_check$diagnostics$slack_used > 0
```

**Solutions:**
1. Review decisions with patient (preferred)
2. Reduce `eps_strict` slightly (1e-4) for more tolerance
3. Keep slack enabled (default) - it finds closest feasible solution

---

### Scenario 2: High Uncertainty

**Problem:** Bootstrap shows very wide confidence intervals

```{r high_uncertainty, eval=FALSE}
boot <- engine_bootstrap(eng, B = 200)

# Wide CIs indicate uncertainty
plot_importance_ci(boot)
```

**Solutions:**
1. Collect more decisions (increase question budget)
2. Check if `tau_equal` is too large (allowing too much indifference)
3. Use sensitivity analysis to confirm robustness

---

### Scenario 3: Degenerate Solutions

**Problem:** Some criteria get zero importance

```{r degenerate, eval=FALSE}
# Check importance
print(eng$importance)
# If any criterion has 0% importance but should matter...
```

**Solutions:**
1. Increase `epsilon_monotone` to 1e-4 (forces level differences)
2. Check if patient actually doesn't care about that criterion
3. Verify levels are meaningfully different to patient

---

## Parameter Interactions

The three parameters interact:

**Rule of thumb:**
```r
eps_strict ≈ tau_equal >> epsilon_monotone
```

**Typical configurations:**

```{r configurations}
# Conservative (stricter constraints)
conservative <- list(
    eps_strict = 0.01,
    tau_equal = 0.01,
    epsilon_monotone = 1e-4
)

# Default (balanced)
default <- list(
    eps_strict = 1e-3,
    tau_equal = 1e-3,
    epsilon_monotone = 1e-6
)

# Permissive (more tolerant)
permissive <- list(
    eps_strict = 1e-4,
    tau_equal = 1e-4,
    epsilon_monotone = 1e-6
)
```

---

## Computational Implications

**Parameter size affects:**

1. **Feasibility:** Smaller values → easier to satisfy constraints
2. **Differentiation:** Larger values → stronger preferences expressed
3. **Numerical stability:** Very small values risk floating-point precision issues

**Recommended ranges:**

| Parameter | Minimum | Default | Maximum |
|-----------|---------|---------|---------|
| `eps_strict` | 1e-6 | 1e-3 | 0.1 |
| `tau_equal` | 1e-6 | 1e-3 | 0.1 |
| `epsilon_monotone` | 1e-7 | 1e-6 | 1e-3 |

---

## Testing Your Configuration

```{r testing_config}
# Test a custom configuration
test_config <- list(
    eps_strict = 0.005,
    tau_equal = 0.005,
    epsilon_monotone = 1e-5
)

eng_test <- engine_create(domains, settings = test_config)

# Collect decisions and test
# ... elicitation ...
# eng_test <- engine_compute(eng_test)

# Run sensitivity analysis to verify robustness
# sens_test <- sensitivity_analysis(eng_test, ...)
```

---

## Key Takeaways

1. **Start with defaults** - they work for 95% of applications
2. **Use sensitivity analysis** - verify results are robust
3. **Adjust conservatively** - change one parameter at a time
4. **Document changes** - note why you adjusted parameters
5. **Validate clinically** - ensure results make clinical sense

**For most healthcare SDM applications, you can ignore these parameters entirely!**

---

## References

- Nocedal & Wright (2006). **Numerical Optimization**. Springer.
- Boyd & Vandenberghe (2004). **Convex Optimization**. Cambridge.

```{r session_info}
sessionInfo()
```
